HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-10-14 docs/HLL_exp_bad/main.tex
bringing in SipHash and updating related work

To complement our attack results, we provide a formal security analysis of HLL. We show that in the ``shadow device'' attack scenario of~\cite{hllvuln}, no security is possible even if the HLL is secretly keyed. On the other hand, we also show that in the most powerful attack setting introduced here, where the attacker has full access to the internals of the HLL algorithm (except for a secret key!), it is possible to provably secure HLL against cardinality manipulation by the simple expedient of replacing its internal hash function by a keyed, variable-input-length pseudo-random function (VIL-PRF). We describe a suitable instantiation of the required VIL-PRF using off-the-shelf cryptographic components.

@@

Desfontaines \emph{et al.}~\cite{cardestprivacy} studied the privacy properties of HLL in two different attack scenarios, referred to as \emph{insider} and \emph{external}. The attack target of~\cite{cardestprivacy} is different from ours, being concerned with loss of privacy, whereas we care about manipulating cardinality estimation. It is argued in~\cite{cardestprivacy} that HLL can be used by organisations to store and process location data, which is inherently sensitive, so losing privacy would leak users' locations and potentially cause great harm. In Section~\ref{sec:attacks} we will discuss the two attack scenarios from~\cite{cardestprivacy} in more detail. 
%In particular, we will show that an adversary who can insert items into the HLL and who can access the HLL cardinality estimate (via typical APIs in HLL implementations) as permitted in the external scenario of~\cite{cardestprivacy} can actually realise the same attacks as the insider adversary can, at least insofar as cardinality estimate manipulation is concerned.

In~\cite{hllvuln}, Reviriego and Ting exploit a vulnerability of HLL to manipulate the cardinality estimate, leading to a five-fold reduction in the estimate compared to the true cardinality. We consider the attack model used in~\cite{hllvuln} to be quite artificial, since it gives the adversary the ability to test whether inserting an item \emph{would} increase the cardinality estimate, without actually having to insert the item. After some clarifications obtained through personal correspondence with the authors of~\cite{hllvuln}, we understand that such a setup might be possible under certain circumstances, for example where the adversary has access to an identical ``shadow'' device; see further discussion in Section~\ref{sec:attacks}. For completeness, we include analyses of this attack scenario, showing that we can drastically improve on the attacks in~\cite{hllvuln} (e.g.\ we can reduce the HLL cardinality estimate by a factor of almost 1,000 instead of the factor of 5 achieved in~\cite{hllvuln}).

@@

Clayton \emph{et al.}~\cite{CCS:ClaPatShrPS19} provide a provable-security treatment of probabilistic data structures in adversarial environments more generally. They use their formalism to analyze Bloom filters, counting (Bloom) filters and count-min sketch data structures, but not HLL. %It is an interesting open problem to use their framework to develop a formal understanding of how HLL can be made secure in the adversarial setting.

@@

\begin{itemize}
\item[S1:] The adversary does not know the details of the target HLL implementation but can access a shadow copy of the HLL instance via its API. It can use the API to insert elements into this shadow sketch, get its cardinality estimate, and reset it to its native empty state. The adversary does not know the state of the targeted HLL sketch, meaning in particular that it does not have the list of elements previously inserted into it by other users. It can insert items into the target HLL sketch, but is otherwise ``blind''. 
\item[S2:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. It can insert items into the HLL sketch, but is otherwise blind. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time. However, it does not know the individual bucket contents.
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It can access and interact with the sketch via an API provided by the sketch owner, as per scenario S3. 
%Note that, using these capabilities, an S4 adversary can compute a lower bound on the HLL cardinality estimate based on its initial snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

As a proof-of-concept, we implemented our attacks against the HLL implementation from~\cite{clahll}, since it is faithful to the original description of HLL given in~\cite{hll}.\footnote{We will make our attack code available on acceptance.}
%at \href{https://github.com/PizzaWhisperer/HLLVuln}{\url{https://github.com/PizzaWhisperer/HLLVuln}}.} 
%\todo{We need to make sure the repo is publicly accessible but anonymised for submission, if we are going to include it.}
For simplicity, we take ${\cal X}$ to be the set of 4-character long ASCII strings, with $\lvert {\cal X} \rvert = 7,311,616= 2^{22.8}$. We choose $h$ to be the 32-bit Murmur3 hash function from~\cite{murmur3code}.
%(note that the library allows the hash function to be specified; we pick Murmur3 for our attack and consider it reasonable that the adversary should know the hash function in use).
We create an HLL sketch using $n=8$ and hence $m=256$ buckets and, for attacks in the setting of an initially non-empty target HLL sketch, initialise it with 1,000 random items representing honest users' data.

@@

\subsection{Security Proof for HLL-PRF}

@@

\begin{theorem}\label{thm:security-of-HLL-with-PRF}
Suppose ${\cal A}$ is an S4 adversary against HLL instantiated with a keyed function $F: \{0,1\}^k \times {\cal D} \rightarrow  \{0,1\}^{n+\ell}$. Then we can construct from ${\cal A}$ a PRF adversary ${\cal B}$ against $F$ such that:
\[
\mbox{Adv}^{\mbox{\tiny S4-RoI}}_{\mbox{\tiny HLL}}({\cal A}) = \mbox{Adv}^{\mbox{\tiny PRF}}_{F}({\cal B}) \,.
\]
Moreover, if ${\cal A}$ makes $q$ queries to its \texttt{insert($\cdot$)} oracle, then ${\cal B}$ makes $q$ queries to its real or random oracle. Finally, the running time of ${\cal B}$ is (roughly) the same as that of ${\cal A}$.
\end{theorem}

@@

@@

A particularly simple construction for $F$ involves composing a collision-resistant hash function $H: {\cal D} \rightarrow  {\cal E}$ onto some domain ${\cal E}$ with a PRF $G: \{0,1\}^k \times {\cal E} \rightarrow \{0,1\}^{n+\ell}$, viz: $F(K,x) = G(K,H(x))$. The result is a PRF with the required input-output signature~\cite[Section 8.2]{BonehShoup}. As a concrete instantiation, one could take $H$ to be SHA-256 with outputs truncated to 128 bits and $G$ to be AES-128 truncated to ${n+\ell}$ bits (since ${n+\ell} \leq 64 < 128$ in practice, this truncation is always possible).

An alternative is to simply use HMAC-SHA-256 with an appropriately truncated output. HMAC is a PRF under reasonable assumptions on the hash function used in its construction~\cite{Bellare15}.

These constructions would be relatively fast on commodity platforms due to widespread hardware support for SHA-256 and AES operations. It would be a simple exercise to add the above PRF proposals to an existing HLL implementation and benchmark them against, say, Murmur3.  The exact slowdown depends on platform specifics, but based on microbenchmarks can be expected to be in the range 2-fold to 5-fold compared to Murmur3. \todo{Put a hard number here.}

@@

We have given a formal analysis of using a PRF as a countermeasure to our attacks. The use of a keyed function sacrifices HLL mergeability when the keys in the different HLLs are independently generated. It is an open problem to find methods for securing HLL sketches against adversarial inputs whilst retaining some form of mergeability. This may also require the introduction of entirely new data structures that trade increased storage for mergeability and improved security. It is also an open problem to close the performance gap that exists between fast but non-cryptographic hashes used in HLL implementations today and the slower but cryptographically strong VIL-PRF that is needed to obtain security.

>>>>

To complement our attack results, we provide a formal security analysis of HLL. We show that in the ``shadow device'' attack scenario of~\cite{hllvuln}, no security is possible even if the HLL is secretly keyed. On the other hand, we also show that in the most powerful attack setting introduced here, where the attacker has full access to the internals of the HLL algorithm (except for a secret key!), it is possible to provably secure HLL against cardinality manipulation by the simple expedient of replacing its internal hash function by a keyed, variable-input-length pseudo-random function (VIL-PRF). We discuss and benchmark a suitable choice for the required VIL-PRF.

@@

Desfontaines \emph{et al.}~\cite{cardestprivacy} studied the privacy properties of HLL in two different attack scenarios, referred to as \emph{insider} and \emph{external}. The attack target of~\cite{cardestprivacy} is different from ours, being focussed on privacy breaches, whereas we care about manipulating cardinality estimation. 
%It is argued in~\cite{cardestprivacy} that HLL can be used by organisations to store and process location data, which is inherently sensitive, so losing privacy would leak users' locations and potentially cause great harm. 
We discuss the two attack scenarios from~\cite{cardestprivacy} in more detail in Section~\ref{sec:attacks}.

%In particular, we will show that an adversary who can insert items into the HLL and who can access the HLL cardinality estimate (via typical APIs in HLL implementations) as permitted in the external scenario of~\cite{cardestprivacy} can actually realise the same attacks as the insider adversary can, at least insofar as cardinality estimate manipulation is concerned.

In~\cite{hllvuln}, Reviriego and Ting exploit a vulnerability of HLL to manipulate the cardinality estimate, leading to a five-fold reduction in the estimate compared to the true cardinality. We consider the attack model used in~\cite{hllvuln} to be quite artificial, since it gives the adversary the ability to test whether inserting an item \emph{would} increase the cardinality estimate, without actually having to insert the item. Such a setup might be possible under certain circumstances, for example where the adversary has access to an identical ``shadow'' device; see further discussion in Section~\ref{sec:attacks}. For completeness, we include analyses of this attack scenario, showing that we can drastically improve on the attacks in~\cite{hllvuln} (e.g.\ we can reduce the HLL cardinality estimate by a factor of almost 1,000 instead of the factor of 5 achieved in~\cite{hllvuln}).

@@

Our work relates to the broader study of the performance of probabilistic data structures in adversarial environments.
Clayton \emph{et al.}~\cite{CCS:ClaPatShrPS19} recently provided a provable-security treatment of this topic. They used their formalism to analyze Bloom filters, counting (Bloom) filters and count-min sketch data structures, but not HLL. Prior work also focussed on Bloom filters~\cite{NaorY15,GerbetKL15} or flooding of hash tables~\cite{LiptonN93,CrosbyW03}. References~\cite{GerbetKL15,CCS:ClaPatShrPS19} provide comprehensive summaries of prior work in this area.

%It is an interesting open problem to use their framework to develop a formal understanding of how HLL can be made secure in the adversarial setting.

@@

\begin{itemize}
\item[S1:] The adversary does not know the details of the target HLL sketch but can access a shadow copy of the HLL sketch via its API. It can use the API to insert elements into this shadow HLL sketch, get its cardinality estimate, and reset it to its native empty state. The adversary does not know the state of the targeted HLL sketch, meaning in particular that it does not have the list of elements previously inserted into it by other users. It can insert items into the target HLL sketch, but is otherwise ``blind''. 
\item[S2:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. It can insert items into the HLL sketch, but is otherwise blind. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time. However, it does not know the individual bucket contents.
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It can access and interact with the sketch via an API provided by the sketch owner, as per scenario S3. 
%Note that, using these capabilities, an S4 adversary can compute a lower bound on the HLL cardinality estimate based on its initial snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

As a proof-of-concept, we implemented our attacks against the HLL implementation from~\cite{clahll}, since it is faithful to the original description of HLL given in~\cite{hll}.\footnote{We will make our attack code available on acceptance.}
%at \href{https://github.com/PizzaWhisperer/HLLVuln}{\url{https://github.com/PizzaWhisperer/HLLVuln}}.} 
%\todo{We need to make sure the repo is publicly accessible but anonymised for submission, if we are going to include it.}
For simplicity, we take ${\cal X}$ to be the set of 4-character long ASCII strings, with $\lvert {\cal X} \rvert = 7,311,616= 2^{22.8}$. We choose $h$ to be the 32-bit MurmurHash3 function from~\cite{murmur3code}.
%(note that the library allows the hash function to be specified; we pick Murmur3 for our attack and consider it reasonable that the adversary should know the hash function in use).
We create an HLL sketch using $n=8$ and hence $m=256$ buckets and, for attacks in the setting of an initially non-empty target HLL sketch, initialise it with 1,000 random items representing honest users' data.

@@

\subsection{Security Analysis of HLL with a PRF}

@@

\begin{theorem}\label{thm:security-of-HLL-with-PRF}
Suppose ${\cal A}$ is an S4 adversary against HLL instantiated with a keyed function $F: \{0,1\}^k \times {\cal D} \rightarrow  \{0,1\}^{n+\ell}$. Then we can construct from ${\cal A}$ a PRF adversary ${\cal B}$ against $F$ such that:
\[
 \mbox{Adv}^{\mbox{\tiny PRF}}_{F}({\cal B}) = \mbox{Adv}^{\mbox{\tiny S4-RoI}}_{\mbox{\tiny HLL}}({\cal A}) \,.
\]
Moreover, if ${\cal A}$ makes $q$ queries to its \texttt{insert($\cdot$)} oracle, then ${\cal B}$ makes $q$ queries to its real or random oracle. Finally, the running time of ${\cal B}$ is (roughly) the same as that of ${\cal A}$.
\end{theorem}

@@

We want $F$ to be competitive with non-cryptographic hashing on small input sizes, as typical in current HLL implementations and applications (e.g.\ MurmurHash3 applied to an IP header). This makes off-the-shelf approaches like HMAC uncompetitive, since even for the shortest inputs, HMAC would require 4 evaluations of the underlying hash compression function.

SipHash~\cite{AumassonB12} is designed specifically for this use case. It is competitive in speed with non-cryptographic hash functions commonly used in HLL implementations such as MurmurHash3. It has also withstood cryptanalysis well~\cite{DobraunigMS14,XinLSL19}. It has 128-bit outputs that can be truncated to the needs of typical HLL implementations (32 and 64 bits).%and is well supported in libraries such as OpenSSL.

\todo{Add a sentence on SipHash-2-4 performance here. Microbenchmark on a short input suffices.}

@@

%A particularly simple construction for $F$ involves composing a collision-resistant hash function $H: {\cal D} \rightarrow  {\cal E}$ onto some domain ${\cal E}$ with a PRF $G: \{0,1\}^k \times {\cal E} \rightarrow \{0,1\}^{n+\ell}$, viz: $F(K,x) = G(K,H(x))$. The result is a PRF with the required input-output signature~\cite[Section 8.2]{BonehShoup}. As a concrete instantiation, one could take $H$ to be SHA-256 with outputs truncated to 128 bits and $G$ to be AES-128 truncated to ${n+\ell}$ bits (since ${n+\ell} \leq 64 < 128$ in practice, this truncation is always possible).

%An alternative is to simply use HMAC-SHA-256 with an appropriately truncated output. HMAC is a PRF under reasonable assumptions on the hash function used in its construction~\cite{Bellare15}.

%These constructions would be relatively fast on commodity platforms due to widespread hardware support for SHA-256 and AES operations. It would be a simple exercise to add the above PRF proposals to an existing HLL implementation and benchmark them against, say, Murmur3.  The exact slowdown depends on platform specifics, but based on microbenchmarks can be expected to be in the range 2-fold to 5-fold compared to Murmur3. \todo{Put a hard number here.}

@@

We have given a formal analysis of using a PRF as a countermeasure to our attacks. The use of a keyed function sacrifices HLL mergeability when the keys in the different HLLs are independently generated. It is an open problem to find methods for securing HLL sketches against adversarial inputs whilst retaining some form of mergeability. This may also require the introduction of entirely new data structures that trade increased storage for mergeability and improved security. 
%It is also an open problem to close the performance gap that exists between fast but non-cryptographic hashes used in HLL implementations today and the slower but cryptographically strong VIL-PRF that is needed to obtain security.
